---
title: "Statistical Inference Course Project"
author: "Artem M."
date: "November 2, 2017"
output:
  pdf_document: default
  html_document: default
subtitle: 'Part 1: Simulation Exercise'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Overview:

In this task we investigate the exponential distribution and compare it with Central Limit Theorem (CLT).
The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is $\frac{1}{\lambda}$ and the standard deviation is also $\frac{1}{\lambda}$.

We will perform a 1000 simluations, where $\lambda = 0.2$ and number of observations in each simulations will be set to 40 ($n = 40$).

Note: For the code of the figures in this project see the Appendix.

#Simulations:

We use R function **rexp** to generate the 1000 random samples of 40 exponentials. To make the results reproducible we will set random seed equal to 10. We will use matrix __exp_dis__ with size  ${n}\times{1000}$ to store the results of the simulations:

```{r simulation}
set.seed(10)
exp_dis=NULL
for(i in 1:1000) exp_dis<-rbind(exp_dis,rexp(40,0.2))

dim(exp_dis)

```
Our simulation generated a matrix of 1000 rows(samples) by 40 columns(observations).


#Sample Mean versus Theoretical Mean


We calculated the sample mean ($\bar{X}$) for our simulated data: 

```{r}
smeans<-as.data.frame(apply(exp_dis,1,mean))
names(smeans)<-"value"
xbar<-round(mean(smeans[,1]),3)
xbar

```
And then found the theoretical mean of the population ($E[X]=\mu= \frac {1}{0.2}$)

```{r}

ex<-1/0.2
ex

```

Then we constructed a sampled means histogram (see Fig.1 ) that shows that it is centered at `r xbar` (a), which is close to the theoretical population mean 5 (b).

```{r echo=FALSE}
library(ggplot2)
tit1<-substitute(paste("Fig.1: Sample histogram: (a) - ",bar(X),"=",m,"; ", "(b) - ",mu,"=",n),
                                list(m=xbar,n=5))
ggplot(smeans, aes(x=value)) + 
        geom_histogram(binwidth=.3, colour="black", fill="white") +
        labs(title=tit1 )+

        labs(x="Value", y="Frequency") +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_vline(aes(xintercept=xbar), 
               color="red", linetype="dashed", size=1)+
        annotate('text', x = xbar +.1, y = 150, label ="(a)")+
        geom_vline(aes(xintercept=5), 
               color="blue", linetype="dashed", size=.5)+
        annotate('text', x = 5-.1, y = 150, label ="(b)")


```


Comparison of the results of the calculation demonstrates that the sample mean is close to the theoretical mean, as such clearly shows its applicability as a good population mean estimate.


# Variability of the sample  


By CLT $\bar{X} \sim N\left(\mu,\frac {\sigma^2}{n}\right)$, where $\sigma^2$ is the variation of the population  being sampled from and the variance of the sample mean is (1)$Var(\bar{X})=\frac{\sigma^2}{n}$ 

We can find the empirical sampling mean variance via R function var:

```{r}

smvar<-var(smeans$value)
round(smvar,3)

```
To find theoretical mean variance we will use variance equation (1) above and replace $\sigma^2$ with squared standard deviation for the exponential distribution  $\Big(\frac{1}{\lambda}\Big)^2$. So the variance equation will take the form $Var(\bar{X})=\frac{\Big(\frac{1}{\lambda}\Big)^2}{n}$.
Variance value is calculated as follows:

```{r}
tmvar<-(1/.2)^2/40
tmvar

```

We will use standard deviation for graphically demonstrate mean variability ($\sqrt{Var(\bar{X})}$) where __sdemp__ - is empirical standard deviation and __sdter__ - theoretical SD.

```{r}

sdemp<-round(sqrt(smvar),3)
sdemp

sdter<-round(sqrt(tmvar),3)
sdter

```
We see that on Fig.2 below, values of sdemp = `r sdemp` and sdter = `r sdter` are very close, and consequently relevant variances.

```{r echo=FALSE}

library(grid)
sd1 <- sprintf("italic(sdemp) == %.3f",sdemp )
sd2 <- sprintf("italic(sdter) == %.3f",sdter )
tit2<-substitute(paste("Fig.2: Empirical SD: sdemp =",m," vs. theoretical SD sdter=",n),
                                list(m=sdemp,n=sdter))
ggplot(smeans, aes(x=value)) + 
        geom_histogram(binwidth=.3, colour="black", fill="white") +
        labs(title=tit2 )+

        labs(x="Value", y="Frequency") +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_vline(aes(xintercept=mean(value)), color="red", linetype="dashed", size=1)+
        annotate("segment", x=xbar, xend=xbar+sdemp, y=50, yend=50, colour="red",
                 arrow=arrow(ends="both", angle=90, length=unit(.3,"cm")))+
        geom_text(x=xbar+.1, y=55, aes(label=sd1), parse=TRUE, hjust=0)+
        annotate('text', x = xbar+.1, y = 150, label ="(a)")+
        geom_vline(aes(xintercept=5), color="blue", linetype="dashed", size=.5)+
        annotate("segment", x=5, xend=5+sdter, y=40, yend=40, colour="blue",
                 arrow=arrow(ends="both", angle=90, length=unit(.3,"cm")))+
        geom_text(x=5+.1, y=35, aes(label=sd2), parse=TRUE, hjust=0)+
        annotate('text', x = 5-.1, y = 150, label ="(b)")
        

```

# Distribution

To demonstrate that the samlping mean distribution is approximately normal we superimposed a normal curve over the  Sample histogram, see Fig.3. 
To keep the frequency count of the y-axis we scaled the normal curve by the factor of the simulations(1000) and bin width(0.3).
As we see on Fig.3 the original Sample histogram closely follows the normal curve, as such we can conclude that the sampling means distirbution is approximately normal.

``` {r echo=FALSE}

ggplot(smeans, aes(x=value)) + 
        geom_histogram(binwidth=.3, colour="black", fill="white") +
        labs(title="Fig.3: Distribution" )+
        labs(x="Value", y="Frequency") +
        theme(plot.title = element_text(hjust = 0.5))+
        stat_function(fun = function(x) dnorm(x, mean = 5, sd = .791)*1000*.3,
    color = "darkred", size = 1)

```




# Appendix

__Fig.1 code__

```{r eval=FALSE}
tit1<-substitute(
        paste("Fig.1: Sample histogram: (a) - ",bar(X),"=",m,"; ", "(b) - ",mu,"=",n),
                                list(m=xbar,n=5))
gmeans<-ggplot(smeans, aes(x=value)) + 
        geom_histogram(binwidth=.3, colour="black", fill="white") +
        labs(title=tit1 )+

        labs(x="Value", y="Frequency") +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_vline(aes(xintercept=xbar), 
               color="red", linetype="dashed", size=1)+
        annotate('text', x = xbar+.1, y = 150, label ="(a)")+
        geom_vline(aes(xintercept=5), color="blue", linetype="dashed", size=.5)+
        annotate('text', x = 5-.1, y = 150, label ="(b)")
gmeans

```

__Fig.2 code__
```{r eval=FALSE}

library(grid)
sd1 <- sprintf("italic(sdemp) == %.3f",sdemp )
sd2 <- sprintf("italic(sdter) == %.3f",sdter )
tit2<-substitute(paste("Fig.2: Empirical SD: sdemp =",m," vs. theoretical SD sdter=",n),
                                list(m=sdemp,n=sdter))
vars<-ggplot(smeans, aes(x=value)) + 
        geom_histogram(binwidth=.3, colour="black", fill="white") +
        labs(title=tit2 )+

        labs(x="Value", y="Frequency") +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_vline(aes(xintercept=mean(value)), color="red", linetype="dashed", size=1)+
        annotate("segment", x=xbar, xend=xbar+sdemp, y=50, yend=50, colour="red",
                 arrow=arrow(ends="both", angle=90, length=unit(.3,"cm")))+
        geom_text(x=xbar+.1, y=55, aes(label=sd1), parse=TRUE, hjust=0)+
        annotate('text', x = xbar+.1, y = 150, label ="(a)")+
        geom_vline(aes(xintercept=5), color="blue", linetype="dashed", size=.5)+
        annotate("segment", x=5, xend=5+sdter, y=40, yend=40, colour="blue",
                 arrow=arrow(ends="both", angle=90, length=unit(.3,"cm")))+
        geom_text(x=5+.1, y=35, aes(label=sd2), parse=TRUE, hjust=0)+
        annotate('text', x = 5-.1, y = 150, label ="(b)")
vars        

```
__Fig.3 code__
``` {r eval=FALSE}

dist<-ggplot(smeans, aes(x=value)) + 
        geom_histogram(binwidth=.3, colour="black", fill="white") +
        labs(title="Fig.3: Distribution" )+
        labs(x="Value", y="Frequency") +
        theme(plot.title = element_text(hjust = 0.5))+
        stat_function(fun = function(x) dnorm(x, mean = 5, sd = .791)*1000*.3,
    color = "darkred", size = 1)

dist
```

